---
title: "Module 1 - Installing spaceNtime"
author: "Read Barbee"
date: "2024-08-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Welcome!

Welcome to the Space to Event (STE), Time to Event (TTE), and Instantaneous Sampling (IS) portion of the workshop! We will be using the R package spaceNtime (Moeller and Lukacs 2021) for all of these analyses. The following steps will help you install and start using the package.


### Step 1: Make sure the devtools package is installed

Run this code chunk to check if the devtools package is installed. If it isn't this code will install it for you. 

```{r, eval = F}
if (!require("devtools", character.only = TRUE)) {
    install.packages("devtools")
    library("devtools", character.only = TRUE)
  print("devtools is now installed!")
} else if(require("devtools", character.only = TRUE)){
  library("devtools", character.only = TRUE)
  print("devtools is already installed!")
}

```



### Step 2: Install the spaceNtime package

Run this code chunk to install the spaceNtime package from GitHub. 
```{r, eval = F}
if(!require("spaceNtime", character.only = TRUE)){
  devtools::install_github("annam21/spaceNtime", force = T, build_vignettes = T)
}
```



### Step 3: Make sure the package is installed properly 

If the package installed correctly, this code will load it. If not, it will give you an error message and tell you to return to step 2.
```{r, eval = F}
if (require("spaceNtime", character.only = TRUE)) {
    library("spaceNtime", character.only = TRUE)
  print("spaceNtime is installed and ready to use!")
} else if(!require("spaceNtime", character.only = TRUE)){
  print("Something went wrong. Try installing spaceNtime again using step 2.")
}
```

### More help

That's it! We're ready to go. If you want more information about the spaceNtime package, you can run the following chunk:

For help files and vignettes:
```{r}
??spaceNtime
```

\
The GitHub page for the R package is here: 
https://github.com/annam21/spaceNtime

\


